---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# topocast

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

*coarse climate grids at the resolution of your terrain*

**Downscale a coarse raster onto fine terrain by moving-window regression.**

```{r downscale, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 4.3, dpi = 140, fig.alt = "A coarse blocky precipitation grid beside the same field downscaled onto fine terrain, sharing one colour scale."}
library(topocast)
library(terra)

set.seed(1)
fine <- rast(nrows = 180, ncols = 180, xmin = 0, xmax = 18000, ymin = 0, ymax = 18000,
             crs = "EPSG:32632")
xy <- crds(fine)
elev <- setValues(fine, 1300 + 700 * sin(xy[, 1] / 2700) + 500 * cos(xy[, 2] / 2200) +
                          260 * sin(xy[, 1] / 760) + 200 * cos(xy[, 2] / 900))
names(elev) <- "elev"

coarse_prec <- aggregate(elev, fact = 12, fun = "mean")
coarse_prec <- 1100 - 0.24 * coarse_prec +
  setValues(coarse_prec, rnorm(ncell(coarse_prec), 0, 22))
names(coarse_prec) <- "prec"

downscaled <- topocast(prec ~ elev, data = coarse_prec, onto = elev, radius = 5)

rng <- range(c(values(coarse_prec), values(downscaled)), na.rm = TRUE)
pal <- hcl.colors(256, "Spectral")

op <- par(mfrow = c(1, 2), mar = c(0.5, 0.5, 2.4, 0.5), oma = c(0, 0, 0, 3), bg = "white")
plot(coarse_prec, col = pal, range = rng, axes = FALSE, legend = FALSE, mar = NA,
     main = "coarse precipitation  (1.2 km)")
plot(downscaled, col = pal, range = rng, axes = FALSE, mar = NA,
     main = "downscaled onto terrain  (100 m)")
par(op)
```

Give `topocast` a coarse variable and a fine predictor it tracks. In a window
around every cell it learns how the variable depends on the predictor, then
evaluates that local relationship on the fine predictor. A 1 km precipitation
grid and a 100 m elevation model become a 100 m precipitation field.

```r
library(topocast)
library(terra)

names(prec_1km) <- "prec"      # coarse variable, what you want at high resolution
names(dem_100m) <- "elev"      # fine predictor it tracks

prec_100m <- topocast(prec ~ elev, data = prec_1km, onto = dem_100m, radius = 15)
```

`prec_100m` is precipitation on the elevation model's grid. Coarse to fine is one
call: name the response and the predictor in a formula, pass the coarse grid as
`data` and the fine grid as `onto`.

## What goes in, what comes out

- **`data`** is the coarse variable you want at higher resolution, as a
  `SpatRaster`, `Raster*`, or `stars` grid. The layer on the left of the formula
  is the response.
- **`onto`** is the target: a fine grid holding the predictor(s) the variable
  tracks, often a digital elevation model. It can also be a set of station or
  plot points carrying those predictors as attributes.
- **the result** is the response on the geometry of `onto`, in the class of
  `onto`: a raster at the fine resolution for a grid target, a prediction column
  for a point target.

When the only fine layer you have is the predictor itself, as in the example
above, `topocast` derives the coarse predictor from `onto` for you, so a single
coarse climate layer and a DEM are enough to start.

## How it works

The relationship is fit locally, in a square window around every coarse cell,
with summed-area tables: each window fit reduces to four lookups per sufficient
statistic, so a radius of 30 costs the same as a radius of 3. The fitted
intercept and slope grids are resampled to the fine grid and combined with the
fine predictors as `fitted = intercept + sum(slope * predictor)`, so the output
carries the fine-scale structure of the terrain with locally varying
coefficients. This is the regression step behind high-resolution climate
surfaces such as CHELSA (Karger et al. 2017); `topocast` runs it locally and
takes any number of named predictors.

## What's in the box

- **One front door.** `topocast()` takes a formula, the coarse `data`, and the
  fine `onto`, and returns the downscaled response.
- **Any number of predictors.** `prec ~ elev + slope + twi` fits elevation,
  slope, topographic wetness, or any aligned covariate jointly. The formula names
  match layers between `data` and `onto`, so a missing layer is reported by name.
- **Several responses in one call.** `cbind(prec, tmin) ~ elev` downscales
  variables that share the terrain together. The window design is built once and
  solved against each response, so a second response costs little more than the
  first.
- **Constant cost per window.** Summed-area tables keep each per-cell fit at four
  lookups for any radius.
- **Common spatial classes.** `SpatRaster`, `Raster*` (raster), and `stars` grids
  are all accepted; the result comes back in the class of `onto`, or the class
  named by `output`.
- **Prediction at points.** Pass an `sf` or `SpatVector` of stations or plots as
  `onto` and receive a prediction column, with the points carrying the fine
  predictor values.
- **Time series in one call.** Pass a stack of periods as `anomaly` to fit the
  baseline once and carry each period onto it, `"ratio"` for precipitation or
  `"additive"` for temperature.
- **Fit quality on the map.** `diagnostics = TRUE` returns an `r.squared` grid
  showing where the terrain relationship is strong, and `clamp = TRUE` bounds the
  output to the observed range of the coarse response.
- **A matrix engine.** `window_regression()` exposes the kernel for callers who
  hold their data as matrices.

## Installation

```r
# install.packages("pak")
pak::pak("gcol33/topocast")
```

## Usage

Several predictors, matched by name between the two grids:

```r
coarse <- c(prec_1km, elev_1km, twi_1km, slope_1km)
names(coarse) <- c("prec", "elev", "twi", "slope")
terrain <- c(elev_100m, twi_100m, slope_100m)
names(terrain) <- c("elev", "twi", "slope")

prec_100m <- topocast(prec ~ elev + twi + slope, data = coarse, onto = terrain,
                      radius = 15)
```

Several responses that share the terrain, downscaled in one pass and returned as
one layer each:

```r
climate_100m <- topocast(cbind(prec, tmin, tmax) ~ elev, data = coarse,
                         onto = terrain, radius = 15)
```

A monthly series sharing one terrain relationship across periods:

```r
series <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 15,
                   anomaly = prec_monthly_1km, type = "ratio")
```

Downscaling straight to plot locations, returned as a column on the points:

```r
at_plots <- topocast(prec ~ elev, data = coarse, onto = plots_sf, radius = 15)
```

The local coefficient grids, to read the fitted lapse rate:

```r
coefs <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 15,
                  coefficients = TRUE)
```

## Documentation

- [Getting started](https://gillescolling.com/topocast/articles/getting-started.html)
- [How moving-window downscaling works](https://gillescolling.com/topocast/articles/how-it-works.html)
- [Full reference](https://gillescolling.com/topocast/reference/)

## Support

> "Software is like sex: it's better when it's free." — Linus Torvalds

I'm a PhD student who builds R packages in my free time because I believe good
tools should be free and open. I started these projects for my own work and
figured others might find them useful too.

If this package saved you some time, buying me a coffee is a nice way to say
thanks. It helps with my coffee addiction.

[![Buy Me A Coffee](https://img.shields.io/badge/-Buy%20me%20a%20coffee-FFDD00?logo=buymeacoffee&logoColor=black)](https://buymeacoffee.com/gcol33)

## License and citation

MIT. If you use `topocast` in published work, please cite it:

```bibtex
@software{topocast,
  author = {Colling, Gilles},
  title  = {topocast: Moving-Window Regression Downscaling of Raster Data},
  year   = {2026},
  url    = {https://github.com/gcol33/topocast}
}
```
