---
title: "Getting started with topocast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with topocast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dev = "svglite",
  fig.ext = "svg"
)
```

A coarse climate grid tells you the precipitation over a 600 m cell. A digital
elevation model tells you the terrain at 100 m. `topocast` joins the two: it
learns how the coarse variable tracks the terrain, locally, and rewrites the
coarse grid at the resolution of the elevation model. The relationship is fit
once per neighbourhood with a moving-window regression, and the cost does not
grow with the window size.

This vignette downscales a synthetic precipitation field, adds a second terrain
predictor, carries a short time series, and drops down to the matrix engine that
does the work.

```{r libraries}
library(topocast)
library(terra)
```

## A landscape to downscale

We need a fine-resolution terrain and a coarse climate observation. We build the
terrain at 100 m as a smooth surface, then aggregate elevation to a 600 m grid
to stand in for a coarse product. Precipitation is observed only on that coarse
grid.

```{r terrain}
set.seed(1)
fine <- rast(nrows = 120, ncols = 120,
             xmin = 0, xmax = 12000, ymin = 0, ymax = 12000,
             crs = "EPSG:32632")
xy <- crds(fine)
elev_fine <- setValues(fine, 1200 + 600 * sin(xy[, 1] / 2500) +
                               400 * cos(xy[, 2] / 2000))
names(elev_fine) <- "elev"
slope_fine <- terrain(elev_fine, v = "slope", unit = "degrees")
names(slope_fine) <- "slope"
```

The coarse grid is the same landscape at 600 m, six fine cells to a side.

```{r aggregate}
elev_coarse  <- aggregate(elev_fine,  fact = 6, fun = "mean")
slope_coarse <- aggregate(slope_fine, fact = 6, fun = "mean")
```

Precipitation falls with elevation and rises on steeper ground, with noise that
the terrain does not explain. We observe it on the coarse grid only.

```{r climate}
prec_coarse <- 900 - 0.18 * elev_coarse + 4 * slope_coarse +
  setValues(elev_coarse, rnorm(ncell(elev_coarse), 0, 20))
names(prec_coarse) <- "prec"
```

The response and its predictors go into one `SpatRaster`, so they share a grid
by construction. The fine predictors go into a second raster on the target grid.

```{r stacks}
coarse  <- c(prec_coarse, elev_coarse, slope_coarse)
terrain <- c(elev_fine, slope_fine)
coarse
```

## One predictor

A single call downscales the coarse precipitation onto the fine elevation. The
formula names the response and the predictor; both names must be layers of
`coarse`, and the predictor must also be a layer of `onto`.

```{r single}
fine_prec <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 4)
fine_prec
```

The result is a 120 by 120 grid where the coarse input was 20 by 20. Plotting
the two side by side shows where the detail comes from: the coarse field sets
the broad level, the elevation carries the texture.

```{r plot-single, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(prec_coarse, main = "coarse precipitation (600 m)")
plot(fine_prec,   main = "downscaled (100 m)")
```

The `radius` is measured in coarse cells. A radius of 4 means each regression
sees a 9 by 9 window of coarse cells, here a 5.4 km neighbourhood. Inside that
window `topocast` fits one intercept and one slope, then resamples those
coefficients to the fine grid and evaluates `intercept + slope * elev` at every
fine cell.

## Several predictors

Naming more layers adds predictors. Each one is matched by name between `data`
and `onto`, so layer order does not matter and a missing layer is an error
rather than a silent mismatch.

```{r multi}
fine_prec2 <- topocast(prec ~ elev + slope, data = coarse, onto = terrain,
                       radius = 4)
```

A misnamed predictor stops the call and lists what is available, which saves a
debugging session over a quietly wrong raster.

```{r missing, error = TRUE}
topocast(prec ~ elev + aspect, data = coarse, onto = terrain, radius = 4)
```

## Several responses at once

Climate work rarely stops at one variable. Precipitation and temperature share
the same terrain, so they share the moving-window design. Naming them together on
the left-hand side with `cbind()` fits all of them in one pass: the window
cross-products that make up the design are built and factored once, then solved
against each response, so a second response costs little more than the first.

```{r several-responses}
tmin_coarse <- 8 - 0.006 * elev_coarse +
  setValues(elev_coarse, rnorm(ncell(elev_coarse), 0, 0.5))
names(tmin_coarse) <- "tmin"
climate <- c(prec_coarse, tmin_coarse, elev_coarse, slope_coarse)

both <- topocast(cbind(prec, tmin) ~ elev + slope, data = climate, onto = terrain,
                 radius = 4)
both
```

The result carries one layer per response, named for it, and each layer matches
the single-response call. The shared design saves the repeated fitting, not the
numbers. The valid-cell mask is complete-case across the responses: a coarse cell
enters a window only where every response and predictor is finite.

## The one-DEM shortcut

The common case has one coarse climate layer and one fine elevation model, and no
coarse predictor in hand. A predictor named in the formula but absent from `data`
is derived from `onto` by aggregating it to the response grid, so the coarse
elevation does not have to be built by hand.

```{r shortcut}
shortcut <- topocast(prec ~ elev + slope, data = prec_coarse, onto = terrain,
                     radius = 4)
```

Here `data` is the coarse precipitation alone; `elev` and `slope` are aggregated
from `terrain` to the coarse grid with `aggregate`, the cell average by default,
before the fit. The downscaled field reproduces the explicit-predictor result to
resampling tolerance.

```{r shortcut-check}
round(max(abs(values(shortcut) - values(fine_prec2)), na.rm = TRUE), 4)
```

## Mapping the local relationship

The reason to downscale by regression rather than by interpolation is the local
relationship between the response and the terrain. `coefficients = TRUE` returns
that relationship as grids: the fitted layer, the intercept, and one slope per
predictor, all on the fine grid.

```{r coefficients, fig.height = 3.2}
grids <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 4,
                  coefficients = TRUE)
names(grids)

par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(grids[["prec"]], main = "downscaled precipitation")
plot(grids[["elev"]], main = "local slope (mm per m)")
```

The slope layer is the local precipitation lapse rate. Where it is steep,
elevation explains most of the variation in precipitation; where it is near zero,
the terrain carries little signal and the downscaled field stays close to the
coarse input.

## Where the fit is strong

How much of the response the terrain actually explains varies across the map.
`diagnostics = TRUE` returns that as an `r.squared` grid: the coefficient of
determination of each window fit, carried onto the fine grid. High values mark
where the terrain relationship is tight and the downscaling adds real detail; low
values mark where the fitted field rests mostly on the coarse level.

```{r diagnostics, fig.height = 3.6}
fit_quality <- topocast(prec ~ elev + slope, data = coarse, onto = terrain,
                        radius = 4, diagnostics = TRUE)
plot(fit_quality[["r.squared"]], main = "local R-squared", range = c(0, 1))
```

## A time series

Climate downscaling usually means many time steps sharing one terrain
relationship. Passing a stack of coarse periods as `anomaly` fits the baseline
once and carries each period onto it. Use `type = "ratio"` for non-negative
variables such as precipitation and `type = "additive"` for variables such as
temperature.

```{r series}
jan <- prec_coarse * 0.4
jul <- prec_coarse * 1.6
months <- c(jan, jul)
names(months) <- c("jan", "jul")

series <- topocast(prec ~ elev + slope, data = coarse, onto = terrain,
                   radius = 4, anomaly = months, type = "ratio")
nlyr(series)
```

The baseline climatology is downscaled with the regression; each period is then
the fine baseline scaled by that period's coarse anomaly. The ratio path guards
a zero baseline by returning `NA` rather than dividing by zero.

```{r plot-series, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(series[["jan"]], main = "January (100 m)")
plot(series[["jul"]], main = "July (100 m)")
```

## Points and other spatial classes

A common target is a set of locations instead of a full grid: vegetation plots,
weather stations, survey sites. Passing those points as `onto` evaluates the
fitted relationship at each one and returns the points with a prediction column.
The points carry the fine predictors as attributes; here they are read from the
elevation model.

```{r sf-available, include = FALSE}
has_sf <- requireNamespace("sf", quietly = TRUE)
```

```{r points, eval = has_sf}
library(sf)

plots <- st_as_sf(
  data.frame(x = c(2000, 4000, 6000, 8000, 10000, 5000),
             y = c(3000, 8000, 5000, 9000,  2000, 6000)),
  coords = c("x", "y"), crs = "EPSG:32632")

at_terrain <- terra::extract(terrain, vect(plots))
plots$elev  <- at_terrain$elev
plots$slope <- at_terrain$slope

at_plots <- topocast(prec ~ elev + slope, data = coarse, onto = plots, radius = 4)
at_plots
```

The fit at a point is the coarse coefficients interpolated to the location and
applied to the point's predictors, the same regression the grid path evaluates
per cell. Overlaying the points on the downscaled field shows the two agree.

```{r plot-points, fig.height = 3.6, eval = has_sf}
pv <- vect(plots)
plot(fine_prec2, main = "downscaled precipitation, with plot predictions")
plot(pv, add = TRUE, pch = 21, cex = 1.3)
text(crds(pv), labels = round(at_plots$prec), pos = 3, cex = 0.8)
```

The same call returns a plain data frame of coordinates and predictions with
`output = "data.frame"`, for joining back to a plot table.

```{r points-df, eval = has_sf}
topocast(prec ~ elev + slope, data = coarse, onto = plots, radius = 4,
         output = "data.frame")
```

Gridded inputs are not tied to `SpatRaster` either. A `Raster*` object from the
raster package or a `stars` object is accepted for `data` and `onto`, and the
result is returned in the class it was given; `output` requests another. The
class sets only the container for the result, and the numbers are the same.

## The matrix engine

`topocast()` is a terra workflow over a matrix engine, `window_regression()`,
which fits the moving-window regression on plain numeric matrices and returns the
intercept and slope grids. It is useful for testing and for callers who already
hold their data as matrices.

```{r engine}
y <- as.matrix(prec_coarse, wide = TRUE)
x <- list(as.matrix(elev_coarse, wide = TRUE),
          as.matrix(slope_coarse, wide = TRUE))
fit <- window_regression(y, x, radius = 4)
str(fit, max.level = 2)
```

The elevation slope recovers the coefficient we simulated. Over a window that
spans most of the grid the local fit converges on the global value.

```{r engine-recover}
big <- window_regression(y, x, radius = 19)
round(median(big$slope[[1]], na.rm = TRUE), 3)   # simulated -0.18
```

## Practical guidance

A few choices shape the result.

**Radius.** The window sets how local the relationship is. A small radius lets
the slope vary across the map but needs enough valid cells to fit; with two
predictors a window must hold at least three cells, and more for a stable fit.
A large radius approaches a single global regression. Start near the scale over
which you expect the climate-terrain relationship to be stationary, often a few
to a couple dozen coarse cells.

**Resampling method.** The coefficient grids are resampled to the fine grid with
`method`, defaulting to `"cubicspline"`. Cubic spline gives smooth coefficient
surfaces; `"bilinear"` is a safer choice when the coefficients are noisy and you
want to avoid overshoot.

**Degenerate windows.** A cell returns `NA` when its window has fewer valid
cells than the model needs, when a predictor has no spread (`min_variance`), or
when the design is singular. `min_cells` raises the valid-cell requirement above
the bare minimum if you want a margin.

**Collinear predictors.** Over a small extent, precipitation and elevation can be
nearly collinear, which makes the local slope unstable even though the fit
succeeds. This is a property of the data, not a defect; widen the window, drop a
predictor, or inspect the slope grid (via `coefficients = TRUE`) for noise.

**Deriving the coarse predictor.** When a predictor is named in the formula but is
not a layer of `data`, it is aggregated from `onto` to the response grid with
`aggregate`, a cell average by default. Set it to another [terra::resample()]
method if a different aggregation suits the predictor.

**Clamping extreme terrain.** The downscaled field is a regression surface, so
where the fine predictor reaches values the coarse window never saw, the local
slope keeps extrapolating linearly. `clamp = TRUE` bounds the result to the
observed range of the coarse response, a guard against runaway values at extreme
elevation. The `r.squared` grid from `diagnostics = TRUE` shows where such a fit
is weak in the first place.

**What the output conserves.** The downscaled field is a regression surface
evaluated on the fine predictors. Averaging it back to the coarse grid does not
reproduce the coarse input exactly, because the coarse residual is not carried.
The `anomaly` path is what restores fidelity to the coarse values over time: each
period's coarse anomaly is carried verbatim onto the fine baseline.

**Coordinate systems.** `data` and `onto` should share a coordinate reference
system. Two systems with the same EPSG code are treated as equal even when their
WKT strings differ, which is common with cross-source lon/lat data such as a DEM
and a climate product from different providers. A genuine mismatch is an error
that names both systems.

See `vignette("how-it-works", package = "topocast")` for the summed-area-table
method behind the constant per-cell cost.
