---
title: "Getting started with topocast"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with topocast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  dev = "svglite",
  fig.ext = "svg"
)
```

A coarse climate grid tells you the precipitation over a 600 m cell. A digital
elevation model tells you the terrain at 100 m. `topocast` joins the two: it
learns how the coarse variable tracks the terrain, locally, and rewrites the
coarse grid at the resolution of the elevation model. The relationship is fit
once per neighbourhood with a moving-window regression, and the cost does not
grow with the window size.

This vignette downscales a synthetic precipitation field, adds a second terrain
predictor, carries a short time series, and drops down to the matrix engine that
does the work.

```{r libraries}
library(topocast)
library(terra)
```

## A landscape to downscale

We need a fine-resolution terrain and a coarse climate observation. We build the
terrain at 100 m as a smooth surface, then aggregate elevation to a 600 m grid
to stand in for a coarse product. Precipitation is observed only on that coarse
grid.

```{r terrain}
set.seed(1)
fine <- rast(nrows = 120, ncols = 120,
             xmin = 0, xmax = 12000, ymin = 0, ymax = 12000,
             crs = "EPSG:32632")
xy <- crds(fine)
elev_fine <- setValues(fine, 1200 + 600 * sin(xy[, 1] / 2500) +
                               400 * cos(xy[, 2] / 2000))
names(elev_fine) <- "elev"
slope_fine <- terrain(elev_fine, v = "slope", unit = "degrees")
names(slope_fine) <- "slope"
```

The coarse grid is the same landscape at 600 m, six fine cells to a side.

```{r aggregate}
elev_coarse  <- aggregate(elev_fine,  fact = 6, fun = "mean")
slope_coarse <- aggregate(slope_fine, fact = 6, fun = "mean")
```

Precipitation falls with elevation and rises on steeper ground, with noise that
the terrain does not explain. We observe it on the coarse grid only.

```{r climate}
prec_coarse <- 900 - 0.18 * elev_coarse + 4 * slope_coarse +
  setValues(elev_coarse, rnorm(ncell(elev_coarse), 0, 20))
names(prec_coarse) <- "prec"
```

The response and its predictors go into one `SpatRaster`, so they share a grid
by construction. The fine predictors go into a second raster on the target grid.

```{r stacks}
coarse  <- c(prec_coarse, elev_coarse, slope_coarse)
terrain <- c(elev_fine, slope_fine)
coarse
```

## One predictor

A single call downscales the coarse precipitation onto the fine elevation. The
formula names the response and the predictor; both names must be layers of
`coarse`, and the predictor must also be a layer of `onto`.

```{r single}
fine_prec <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 4)
fine_prec
```

The result is a 120 by 120 grid where the coarse input was 20 by 20. Plotting
the two side by side shows where the detail comes from: the coarse field sets
the broad level, the elevation carries the texture.

```{r plot-single, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(prec_coarse, main = "coarse precipitation (600 m)")
plot(fine_prec,   main = "downscaled (100 m)")
```

The `radius` is measured in coarse cells. A radius of 4 means each regression
sees a 9 by 9 window of coarse cells, here a 5.4 km neighbourhood. Inside that
window `topocast` fits one intercept and one slope, then resamples those
coefficients to the fine grid and evaluates `intercept + slope * elev` at every
fine cell.

## Several predictors

Naming more layers adds predictors. Each one is matched by name between `data`
and `onto`, so layer order does not matter and a missing layer is an error
rather than a silent mismatch.

```{r multi}
fine_prec2 <- topocast(prec ~ elev + slope, data = coarse, onto = terrain,
                       radius = 4)
```

A misnamed predictor stops the call and lists what is available, which saves a
debugging session over a quietly wrong raster.

```{r missing, error = TRUE}
topocast(prec ~ elev + aspect, data = coarse, onto = terrain, radius = 4)
```

## The one-DEM shortcut

The common case has one coarse climate layer and one fine elevation model, and no
coarse predictor in hand. A predictor named in the formula but absent from `data`
is derived from `onto` by aggregating it to the response grid, so the coarse
elevation does not have to be built by hand.

```{r shortcut}
shortcut <- topocast(prec ~ elev + slope, data = prec_coarse, onto = terrain,
                     radius = 4)
```

Here `data` is the coarse precipitation alone; `elev` and `slope` are aggregated
from `terrain` to the coarse grid with `aggregate`, the cell average by default,
before the fit. The downscaled field reproduces the explicit-predictor result to
resampling tolerance.

```{r shortcut-check}
round(max(abs(values(shortcut) - values(fine_prec2)), na.rm = TRUE), 4)
```

## Mapping the local relationship

The reason to downscale by regression rather than by interpolation is the local
relationship between the response and the terrain. `coefficients = TRUE` returns
that relationship as grids: the fitted layer, the intercept, and one slope per
predictor, all on the fine grid.

```{r coefficients, fig.height = 3.2}
grids <- topocast(prec ~ elev, data = coarse, onto = terrain, radius = 4,
                  coefficients = TRUE)
names(grids)

par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(grids[["prec"]], main = "downscaled precipitation")
plot(grids[["elev"]], main = "local slope (mm per m)")
```

The slope layer is the local precipitation lapse rate. Where it is steep,
elevation explains most of the variation in precipitation; where it is near zero,
the terrain carries little signal and the downscaled field stays close to the
coarse input.

## A time series

Climate downscaling usually means many time steps sharing one terrain
relationship. Passing a stack of coarse periods as `anomaly` fits the baseline
once and carries each period onto it. Use `type = "ratio"` for non-negative
variables such as precipitation and `type = "additive"` for variables such as
temperature.

```{r series}
jan <- prec_coarse * 0.4
jul <- prec_coarse * 1.6
months <- c(jan, jul)
names(months) <- c("jan", "jul")

series <- topocast(prec ~ elev + slope, data = coarse, onto = terrain,
                   radius = 4, anomaly = months, type = "ratio")
nlyr(series)
```

The baseline climatology is downscaled with the regression; each period is then
the fine baseline scaled by that period's coarse anomaly. The ratio path guards
a zero baseline by returning `NA` rather than dividing by zero.

```{r plot-series, fig.height = 3.2}
par(mfrow = c(1, 2), mar = c(2, 2, 2, 4))
plot(series[["jan"]], main = "January (100 m)")
plot(series[["jul"]], main = "July (100 m)")
```

## The matrix engine

`topocast()` is a terra workflow over a matrix engine, `window_regression()`,
which fits the moving-window regression on plain numeric matrices and returns the
intercept and slope grids. It is useful for testing and for callers who already
hold their data as matrices.

```{r engine}
y <- as.matrix(prec_coarse, wide = TRUE)
x <- list(as.matrix(elev_coarse, wide = TRUE),
          as.matrix(slope_coarse, wide = TRUE))
fit <- window_regression(y, x, radius = 4)
str(fit, max.level = 2)
```

The elevation slope recovers the coefficient we simulated. Over a window that
spans most of the grid the local fit converges on the global value.

```{r engine-recover}
big <- window_regression(y, x, radius = 19)
round(median(big$slope[[1]], na.rm = TRUE), 3)   # simulated -0.18
```

## Practical guidance

A few choices shape the result.

**Radius.** The window sets how local the relationship is. A small radius lets
the slope vary across the map but needs enough valid cells to fit; with two
predictors a window must hold at least three cells, and more for a stable fit.
A large radius approaches a single global regression. Start near the scale over
which you expect the climate-terrain relationship to be stationary, often a few
to a couple dozen coarse cells.

**Resampling method.** The coefficient grids are resampled to the fine grid with
`method`, defaulting to `"cubicspline"`. Cubic spline gives smooth coefficient
surfaces; `"bilinear"` is a safer choice when the coefficients are noisy and you
want to avoid overshoot.

**Degenerate windows.** A cell returns `NA` when its window has fewer valid
cells than the model needs, when a predictor has no spread (`min_variance`), or
when the design is singular. `min_cells` raises the valid-cell requirement above
the bare minimum if you want a margin.

**Collinear predictors.** Over a small extent, precipitation and elevation can be
nearly collinear, which makes the local slope unstable even though the fit
succeeds. This is a property of the data, not a defect; widen the window, drop a
predictor, or inspect the slope grid (via `coefficients = TRUE`) for noise.

**Deriving the coarse predictor.** When a predictor is named in the formula but is
not a layer of `data`, it is aggregated from `onto` to the response grid with
`aggregate`, a cell average by default. Set it to another [terra::resample()]
method if a different aggregation suits the predictor.

**What the output conserves.** The downscaled field is a regression surface
evaluated on the fine predictors. Averaging it back to the coarse grid does not
reproduce the coarse input exactly, because the coarse residual is not carried.
The `anomaly` path is what restores fidelity to the coarse values over time: each
period's coarse anomaly is carried verbatim onto the fine baseline.

**Coordinate systems.** `data` and `onto` should share a coordinate reference
system. Two systems with the same EPSG code are treated as equal even when their
WKT strings differ, which is common with cross-source lon/lat data such as a DEM
and a climate product from different providers. A genuine mismatch is an error
that names both systems.

See `vignette("how-it-works", package = "topocast")` for the summed-area-table
method behind the constant per-cell cost.
